---
title: "9: Data Visualization"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## LESSON OBJECTIVES
1. Perform simple data visualizations in the R package `ggplot`
2. Develop skills to adjust aesthetics and layers in graphs
3. Apply a decision tree framework for appropriate graphing methods

## SET UP YOUR DATA ANALYSIS SESSION

```{r, message = FALSE, warning = FALSE}
getwd()
library(tidyverse)

PeterPaul.chem.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
PeterPaul.nutrients.gathered <- read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")
PeterPaul.chem.nutrients.summaries <- read.csv("./Data/Processed/NTL-LTER_Lake_Summaries_PeterPaul_Processed.csv")
EPAair <- read.csv("./Data/Processed/EPAair_O3_PM25_NC1718_processed.csv")

EPAair$Date <- as.Date(EPAair$Date, format = "%Y-%m-%d")
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")

```

## GGPLOT

ggplot, called from the package `ggplot2`, is a graphing and image generation tool in R. This package is part of tidyverse. While base R has graphing capabilities, ggplot has the capacity for a wider range and more sophisticated options for graphing. 
ggplot has only a few rules: 

* The first line of ggplot code always starts with `ggplot()`
* A data frame must be specified within the `ggplot()` function. Additional datasets can be specified in subsequent layers.
* Aesthetics must be specified, most commonly x and y variables but including others. Aesthetics can be specified in the `ggplot()` function or in subsequent layers.
* Additional layers must be specified to fill the plot. 

### Geoms

Here are some commonly used layers for plotting in ggplot: 

* geom_bar
* geom_histogram
* geom_freqpoly
* geom_boxplot
* geom_violin
* geom_dotplot
* geom_point
* geom_errorbar
* geom_smooth
* geom_line
* geom_area
* geom_abline (plus geom_hline and geom_vline)
* geom_text

### Aesthetics
Here are some commonly used aesthetic types that can be manipulated in ggplot: 

* color
* fill
* shape
* size
* transparency

### Plotting continuous variables over time: Scatterplot
```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
ggplot(EPAair, aes(x = Date, y = Ozone)) + 
  geom_point()

O3plot <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone))
print(O3plot)

# Fix this code
O3plot2 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone), color = "blue") #result in blue plots
print(O3plot2)

O3plot3 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone, color = "blue")) #result in other color but "blue" word in legend. Aesthetics aes() specifies data, so you have to specify color outside aes function
print(O3plot3)

# Add additional variables
PMplot <- 
  ggplot(EPAair, aes(x = month, y = PM2.5, shape = as.factor(year), color = Site.Name)) +
  geom_point()
print(PMplot)
str(EPAair$month)

# Separate plot with facets
PMplot.faceted <-
  ggplot(EPAair, aes(x = month, y = PM2.5, shape = as.factor(year))) +
  geom_point() +
  facet_wrap(vars(Site.Name), nrow = 3) #without specifying nrow, R will squeeze everything into one same row
print(PMplot.faceted)

PMplot.faceted1 <-
  ggplot(EPAair, aes(x = month, y = PM2.5, shape = as.factor(year))) +
  geom_point() +
  facet_wrap(vars(Site.Name, as.factor(year)), nrow = 3) #without specifying nrow, R will squeeze everything into one same row
print(PMplot.faceted1)

PMplot.faceted2 <-
  ggplot(EPAair, aes(x = month, y = PM2.5)) +
  geom_point() +
  facet_grid(Site.Name ~ year)
print(PMplot.faceted2)

PMplot.faceted3 <-
  ggplot(EPAair, aes(x = month, y = Ozone)) +
  geom_point() +
  facet_grid(Site.Name ~ year)
print(PMplot.faceted3)

PMplot.faceted4 <-
  ggplot(EPAair, aes(x = month, y = Ozone)) +
  geom_point() +
  facet_grid(EPAair$year) #possible to just have one variable specified in facet_grid
print(PMplot.faceted4)

# Filter dataset within plot building
O3plot.Blackstone <- 
  ggplot(subset(EPAair, Site.Name == "Blackstone"), aes(x = Date, y = Ozone)) + 
  geom_point() +
  geom_line()
print(O3plot.Blackstone)

# Exercise: build your own scatterplots of PeterPaul.chem.nutrients

# 1. 
# Plot surface temperatures by day of  year. 
# Color your points by year, and facet by lake in two rows.
ggplot(subset(PeterPaul.chem.nutrients, depth=0), aes(x=daynum, y=temperature_C, color=year4)) +
  geom_point(size=0.1)+
  facet_wrap(vars(lakename), nrow=2)

#2. 
# Plot temperature by date. Color your points by depth.
# Change the size of your point to 0.5
ggplot(PeterPaul.chem.nutrients, aes(x=sampledate, y=temperature_C, color=depth)) + geom_point(size=0.5)

```
### Plotting the relationship between two continuous variables: Scatterplot

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
lightvsDO <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point()
print(lightvsDO)

# Adjust axes
lightvsDOfixed <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point() +
  xlim(0, 250) +
  ylim(0, 20)
print(lightvsDOfixed)

# Depth in the fields of limnology and oceanography is on a reverse scale
tempvsdepth <- 
  #ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth)) +
  ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse() #reverse y axis
print(tempvsdepth)

NvsP <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = tn_ug, color = depth)) +
  geom_point() +
  geom_smooth(method = lm) + #If we don't put anything, the smooth line will be curvy. The grey band around smooth line is the confidence interval
  geom_abline(aes(slope = 16, intercept = 0))
print(NvsP) #The grey points on the plot means there are points with no measurement on one of the field (nitrogen or phosphorus)
#In the plot we can see more points above the abline. This means the lake system is more nitrogen rich than phosphorus

# Exercise: Plot relationships between air quality measurements

# 1. 
# Plot AQI values for ozone by PM2.5, colored by site. 
# Add a line of best fit for the linear regression of these variables.
ggplot(EPAair, aes(x=PM2.5,y=Ozone, col=Site.Name)) +
  geom_point() +
  geom_smooth(method = lm) #If put aes in geom_point, geom_smooth will say missing values. We can also add the aes argument in geom_smooth()

ggplot(EPAair, aes(x=PM2.5,y=Ozone)) +
  geom_point(aes(col=Site.Name)) +
  geom_smooth(method = lm) 
  
```

### Plotting continuous vs. categorical variables

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
 # Barplot + error bars
PeterPaul.nutrient.summaries <- PeterPaul.nutrients.gathered %>%
  group_by(lakename, nutrient) %>%
  summarise(sd = sd(concentration), 
            mean = mean(concentration))

Nutrientplot <- 
  ggplot(PeterPaul.nutrients.gathered) +
  geom_bar(aes(x = lakename, y = concentration, fill = as.factor(nutrient)), # why did we use fill? Because the bar has both border and fill. If you don't specify fill, they will look like grey blocks with colored borders
           position = "dodge", stat = "summary", fun.y = "mean")             # what's happening here?
print(Nutrientplot)

Nutrientplot2 <- 
  ggplot(PeterPaul.nutrient.summaries, aes(x = lakename, y = mean, fill = as.factor(nutrient))) + 
  geom_bar(stat = "identity", position = "dodge") + # what does the stat command do? Because in the summary file we already have the mean calculated, we just need to use stat = identity instead of summary
  geom_errorbar(aes(ymin = mean-sd, ymax = mean+sd), # how do we specify error bars?
                 position = "dodge")
print(Nutrientplot2)

# Are there more effective ways to produce summary stats for categories?

# Box and whiskers plot
Nutrientplot3 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_boxplot(aes(color = nutrient)) # Why didn't we use "fill"? because the data are still points not bar?
print(Nutrientplot3)

# Dot plot
Nutrientplot4 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_dotplot(aes(color = nutrient), binaxis = "y", binwidth = 1, 
               stackdir = "center", position = "dodge") #
print(Nutrientplot4)

# Violin plot
Nutrientplot5 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_violin(aes(color = nutrient)) #
print(Nutrientplot5)

# Frequency polygons
# Using a tidy dataset
Nutrientplot6 <-
  ggplot(PeterPaul.chem.nutrients) +
  geom_freqpoly(aes(x = tn_ug), color = "black") +
  geom_freqpoly(aes(x = tp_ug), color = "darkblue") +
  geom_freqpoly(aes(x = nh34), color = "darkgray") +
  geom_freqpoly(aes(x = no23), color = "gray") +
  geom_freqpoly(aes(x = po4), color = "blue") 
print(Nutrientplot6)

# Using a gathered dataset
Nutrientplot7 <-   
  ggplot(PeterPaul.nutrients.gathered) +
  geom_freqpoly(aes(x = concentration, color = nutrient))
print(Nutrientplot7)

# Exercise: Plot distributions of AQI values for EPAair

# 1. 
# Create a bar chart plus standard deviation error bars for PM2.5, divided by year. 
# Create separate bars for each site. 
EPAair.summaries <- EPAair %>%
  group_by(Site.Name, year) %>%
  summarise(
    mean = mean(PM2.5, na.rm=TRUE),
    sd = sd(PM2.5, na.rm=TRUE))
str(EPAair.summaries)
ggplot(EPAair.summaries, aes(x=as.factor(year), y=mean, fill=Site.Name))+
  geom_bar(position="dodge", stat="identity") +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),# how do we specify error bars?
                 position = "dodge")


# 2. 
# Create a new plot that better depicts the distribution of PM2.5 concentrations. 
# Divide your graph by year and site.

```
