---
title: "Assignment 6: Generalized Linear Models"
author: "Ying Wei Jong"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics (ENV872L) on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Use the lesson as a guide. It contains code that can be modified to complete the assignment.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
Space for your answers is provided in this document and is indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio. 
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
You will need to have the correct software installed to do this (see Software Installation Guide)
Press the `Knit` button in the RStudio scripting panel.
This will save the PDF output in your Assignments folder.
8. After Knitting, please submit the completed exercise (PDF file) to the dropbox in Sakai. Please add your last name into the file name (e.g., "Salk_A06_GLMs.pdf") prior to submission.

The completed exercise is due on Tuesday, 26 February, 2019 before class begins.

## Set up your session 
1. Set up your session. Upload the EPA Ecotox dataset for Neonicotinoids and the NTL-LTER raw data file for chemistry/physics. 

2. Build a ggplot theme and set it as your default theme.

```{r}
#1
EPA_Ecotox <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Mortality_raw.csv", header=T)
LTER <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", header=T)

#2
library(ggplot2)
my.theme <- theme_bw(base_size = 12) + 
  theme(axis.text=element_text(color="gray0"), legend.position = "right")
theme_set(my.theme)
```

## Neonicotinoids test
Research question: Were studies on various neonicotinoid chemicals conducted in different years? 

3. Generate a line of code to determine how many different chemicals are listed in the Chemical.Name column.

4. Are the publication years associated with each chemical well-approximated by a normal distribution? Run the appropriate test and also generate a frequency polygon to illustrate the distribution of counts for each year, divided by chemical name. Bonus points if you can generate the results of your test from a pipe function. No need to make this graph pretty.

5. Is there equal variance among the publication years for each chemical? Hint: var.test is not the correct function.

```{r}
#3
unique(EPA_Ecotox$Chemical.Name)

#4
YearForEachChem <- by(EPA_Ecotox, EPA_Ecotox$Chemical.Name, function(x) shapiro.test(x$Pub..Year))
sapply(YearForEachChem, print)
#For each chemical, the publication years associated with it is not well-approximated by a normal distribution

#Pipe version didnt work....
library(tidyverse)
YearForEachChem2 <- EPA_Ecotox %>%
  group_by(Chemical.Name) %>%
  shapiro.test(Pub..Year) #I get an error message: unused argument. Taylor help!

ggplot(EPA_Ecotox, aes(x=Pub..Year, col = Chemical.Name))+
  geom_freqpoly()

#5
YearForEachChem3 <- by(EPA_Ecotox, EPA_Ecotox$Chemical.Name, function(x) var(x$Pub..Year))
sapply(YearForEachChem3, print) #They do look like they have non-equal variance
#install.packages("car")
library(car)
leveneTest(Pub..Year ~ Chemical.Name, data = EPA_Ecotox) #we reject the hypothesis that the groups have equal variance

```

6. Based on your results, which test would you choose to run to answer your research question?

> ANSWER: One-way ANOVA. Continuous response with one categorical explanatory variable with more than two categories

7. Run this test below. 

8. Generate a boxplot representing the range of publication years for each chemical. Adjust your graph to make it pretty.
```{r}
#7
lm1 <- lm(Pub..Year ~ Chemical.Name, data = EPA_Ecotox)
summary(lm1)

#8
ggplot(EPA_Ecotox,aes(y=Pub..Year, x=Chemical.Name)) +
  geom_boxplot(aes(fill=Chemical.Name)) +
  xlab("Chemicals") +
  ylab("Published Year")+
  ggtitle("Publised Year Associated With Each Chemical")+
  theme(legend.position = "none")
  
```


9. How would you summarize the conclusion of your analysis? Include a sentence summarizing your findings and include the results of your test in parentheses at the end of the sentence. 

> ANSWER: The studies on various neonicotinoid chemicals were conducted in different years (One-way ANOVA; F=33.21, df=1274, p<0.0001). 

## NTL-LTER test
Research question: What is the best set of predictors for lake temperatures in July across the monitoring period at the North Temperate Lakes LTER? 

11. Wrangle your NTL-LTER dataset with a pipe function so that it contains only the following criteria: 

* Only dates in July (hint: use the daynum column). No need to consider leap years.
* Only the columns: lakename, year4, daynum, depth, temperature_C
* Only complete cases (i.e., remove NAs)

12. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature. Run a multiple regression on the recommended set of variables. 

```{r}
#11
LTER2 <- LTER %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  filter(daynum %in% c(182:212)) %>%
  na.omit(lakename, year4, daynum, depth, temperature_C)
  
#12
LTERAIC <- lm(temperature_C ~ year4+daynum+depth,data=LTER2)
step(LTERAIC) #It seems to be best to omit nothing
lm2 <- lm(temperature_C ~ year4+daynum+depth,data=LTER2)
summary(lm2)
```

13. What is the final linear equation to predict temperature from your multiple regression? How much of the observed variance does this model explain?

> ANSWER: temperature = -6.45 + 0.01*year4 + 0.041*daynum - 1.95*depth. This model explains 74.2% of the observed variance

14. Run an interaction effects ANCOVA to predict temperature based on depth and lakename from the same wrangled dataset.

```{r}
#14
lm3 <- lm(temperature_C ~ depth * lakename, data=LTER2)
summary(lm3)
```

15. Is there an interaction between depth and lakename? How much variance in the temperature observations does this explain?

> ANSWER: Yes, most combinations of depth and lakename have significant impacts on temperature observations. This model explains about 78.6% of the variance in temperature observations.

16. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r}
#16
library(viridis)
library(RColorBrewer)
ggplot(LTER2, aes(x=depth, y=temperature_C, col=lakename)) +
  geom_point()+
  scale_y_reverse()+
  scale_color_brewer(palette = "YlGnBu")
```

